Dark matter halo shapes in the Auriga simulations

### 2019-08-21 Document/paper.tex
edit

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{correlations_twisting_triaxiality_MHD.pdf}
\end{center}
\caption{Twisting in the halo-disc alignment (measured as the standard deviation of the alignment angles)
  as a function of the halo triaxialiy at the radius indicated
  in each panel's title.
  The label with the $\rho$ value corresponds to the Spearman's rank
  correlation coefficient (mean value and uncertainty estimated via
  jackknife resampling) while the line shows the result of a
  minimum squares fit to the data.
  Twisting does not correlate with triaxiality in the same way at every
  radius. The strongest correlation appears around $0.12\, R_{200}$ ($28\pm2$ kpc).} 
\label{fig:alignment_correlations}
\end{figure*}

@@

The left panel of Figure \ref{fig:triaxial_cumulative} shows that in
the DMO sample the median triaxiality is larger than $2/3$ (a
typical value that marks the transition from triaxiality to
prolateness) at all radii. Furthermore this median value increases 
as we move towards the inner part of the halo.
The middle panel in the same Figure~\ref{fig:triaxial_cumulative}
shows how the trends in the MHD simulations go 
in the opposite direction with respect to the DMO results.
There the median triaxiality is always smaller than $1/3$ (a typical
value marking the transition from triaxiality to prolateness) and this
median value decreases as we move towards the inner part of the halo.

@@

In six halos the alignment changes significantly as a
function of radius,  
as  can be seen by the twisting in the shape ellipsoids. 
Because we are dealing with oblate halos $b/a\approx 1$, note that the
directions corresponding to the medium and major axis could be subject
to noise. 
For this reason, and in order to better quantify the twisting for each
halo, we take the standard deviation of the angle with the minor
axis, which should be less susceptible to noise.
With this measurement, twisting becomes a global property of the halo,
as the standard deviation is measured over all the different radii at
which we have quantified the alignment.

Figure~\ref{fig:alignment_correlations} shows the correlation between
the twisting and the corresponding triaxiality at the four different
radii at which the shape directions are measured.
The median, mean and standard deviation of the twisting 
are $2$, $6$ and $8$ degrees, respectively.
The correlation between twisting and triaxiality also changes with
radius as measured by the Spearman's rank correlation coefficient.
The largest correlation is obtained for the halo triaxiality measured at
$0.12R_{200}$. 
The most important consequence of this changing degree in correlation
is that twisting cannot be explained only by large triaxiality values
and a possibly noisy determination of the shape axis.

@@

The presence of baryons produces rounder dark matter halos than found for the 
corresponding DMO \textbf{}counterpart. To better understand this relationship we examine the correlation
between halo shape and baryonic disc properties.  
Looking into the measurements already reported by \cite{auriga} and
\cite{Pakmor17}, we find three baryonic quantities that have the
strongest correlation with DM halo triaxiality: the central gas
density in a sphere of radius $1$ kpc, the disc to total mass ratio, and
the optical radius.

@@

In some cases, the halo shape also twists in the presence of baryons, as opposed to
the situation in DMO simulations where the shape is coherent as a
function of radius. In Figure~\ref{fig:twisting_correlations}, we show the correlation
between twisting and the same baryonic properties that showed a high
degree of correlation with the triaxiality.
In this case the disc to total ratio does not have a significant
degree of correlation. Instead, the disc optical radius and the gas density
at the core are the most correlated. 
The correlation is such that extended stellar discs with a very low
gas density at their cores show the most coherent halo shapes.

Recently, \citet{2017MNRAS.469..594B} reported that MW-like DM haloes can
be roughly split into two distinct families: stalled and accreting.
Stalled haloes assembled early and have remained almost unchanged
in the last $7$ Gyr, while accreting halos have been growing
constantly. \citet{2017MNRAS.469..594B} also link the two populations to
differences in the dark matter filaments feeding the halos. 
Stalled halos at $z=0$ are embedded in a single large thick filament, while
accreting halos are pierced by many thinner filaments. 
To what extent can the twisting and high triaxiality present in the MHD
halos be linked to this dicotomy?

In Figure~\ref{fig:twisting_age} we show the correlation between the
twisting and the age of the bulge, the age of the stellar disc, and the
DM halo formation redshift. The ages are estimated directly as the mean 
value in the stellar age distribution for each component, while the formation redshift is
defined as the redshift at which the halo has for the first time surpassed half of its final $z=0$
mass. In the three cases we find a correlation between age and twisting
that goes in the direction claimed by \citet{2017MNRAS.469..594B}:
components with early assembly (i.e. stalled halos) have a smaller degree of
twisting. A detailed study to interpret this with respect to the cosmic 
web location of the halos is however beyond the scope of this paper.

@@

% JEFR I've commented this part because the plots are showing the 
% alignment with the disc angular momentum not shape. 
% This part has to be adjusted accordingly.
%\cite{JingSuto02} reported a shell alignment twisting in their dark
%matter only simulations as measured in the direction changes of large
%and medium shape axis. 
%They only had three high resolution MW-like halos (with
%$\sim 10^{6}$ inside the virial radius) and could not make a
%statistical statement about the impact of this effect.
%However, they see the twisting as an artifact resulting from high values %of the
%$b/a$ ratios that make the determination of the medium axis noisy.
%For that reason we measured the twisting using the minor axis
%alignments, however they do not report these alignments.

%\cite{Bailin05} used seven hydrodynamic simulations studied the disc
%halo alignment as a function of radius.
%For radii smaller than $0.1R_{200}$ they find a strong alignment
%between the two minor axis of the halo and the disc. 
%For larger radii the angle between the components changes showing a
%weak correlation between the two components. 
%\cite{Debattista13} performed a study on the disc-halo alignment
%performing controlled experiments on three different halos to study
%changes the evolutionary change that drive the minor axis in the disc
%and the halo to be aligned, their study by construction focuses only
%on radii around $0.1R_{200}$.
%Nevertheless, similarly to \cite{Bailin05}, they also report a strong
%alignment in the region where the disc dominated. 
%In conclusion, the twisting in the dark matter halo shapes seems to be
%a feature already reported in the literature with smaller samples than
%the one we use here. 
%The strongest point of depart is that in our case the alignments do
%not grow stronger as one approached the disc.
%It is very likely that the detailed radial dependence of the twisting
%depends on the hydrodynamics and feedback implementations.

We now use the results reported by \cite{LM10} and \cite{Bovy16}
to place our results in an observational context.
\cite{LM10} used observations of the Sagittarius tidal stream to
constrain the shape of the gravitational potential.
They point out that previous studies that assumed an
axisymetric galactic potential were not able to fit all the available
dynamic constraints for the Sagittarius stream, motivating 
their approach of using a rigid triaxial potential with coaxial potential
ellipsoids for the dark matter component.  
Their results constrain the triaxiality of this potential
component.  They also translate their results into a triaxiality of the density
contours (which could be compared against our results)
 to be $c/a=0.44$ and $b/a=0.97$ at a radius of $\sim 40$~kpc. 
They do not report any uncertainties for these two values. 
Looking at their plots for the quality of fit criterion as a function
of dark halo axial scales (their Figure 5), we choose a conservative $10\%$
relative uncertainty. One surprising element in their results is that 
the major axis of the
halo shape is perpendicular to the stellar disc plane.

@@

\section*{Acknowledgements}
This project has received funding from the European Union's Horizon
2020 Research and Innovation Programme under the Marie
Sk\l{}odowska-Curie grant agreement No 734374.

>>>>

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{correlations_twisting_triaxiality_MHD.pdf}
\end{center}
\caption{Twisting in the halo-disc alignment (measured as the standard
  deviation of the alignment angles at four different radii)
  as a function of the halo triaxialiy at the radius indicated
  in each panel's title.
  The label with the $\rho$ value corresponds to the Spearman's rank
  correlation coefficient (mean value and uncertainty estimated via
  jackknife resampling) while the line shows the result of a
  minimum squares fit to the data.
  Twisting does not correlate with triaxiality in the same way at every
  radius. The strongest correlation appears around $0.12\, R_{200}$ ($28\pm2$ kpc).} 
\label{fig:alignment_correlations}
\end{figure*}

@@

The left panel of Figure \ref{fig:triaxial_cumulative} shows that in
the DMO sample the median triaxiality is larger than $2/3$ (a
typical value that marks the transition from triaxiality to
prolateness) at all radii. Furthermore this median value increases 
as we move towards the inner part of the halo.
The middle panel in the same Figure~\ref{fig:triaxial_cumulative}
shows how the trends in the MHD simulations go 
in the opposite direction with respect to the DMO results.
There the median triaxiality is always smaller than $1/3$ (a typical
value marking the transition from oblateness to triaxiality) and this
median value decreases as we move towards the inner part of the halo.

@@

In six halos the alignment changes significantly as a
function of radius produced by the twisting in the shape ellipsoids. 
In contrast, the DMO simulations show coherent halo shapes as a
function of radius without twisting. 
Because we are dealing with oblate halos $b/a\approx 1$, note that the
directions corresponding to the medium and major axis could be subject
to noise. 
For this reason, and in order to better quantify the twisting for each
halo, we take the standard deviation of the angle with the minor
axis, which should be less susceptible to noise.
With this measurement, twisting becomes a global property of the halo,
as the standard deviation is measured over all the different radii at
which we have quantified the alignment.

Figure~\ref{fig:alignment_correlations} shows the correlation between
the twisting and the corresponding triaxiality at the four different
radii at which the shape directions are measured.
The median, mean and standard deviation of the twisting 
are $2$, $6$ and $8$ degrees, respectively.
The correlation between twisting and triaxiality also changes with
radius as measured by the Spearman's rank correlation coefficient.
The largest correlation is obtained for the halo triaxiality measured at
$0.12R_{200}$. 
The most important consequence of this changing degree of positive
correlation is that twisting cannot be explained only by small
triaxiality (large asphericity) values and a possibly noisy
determination of the shape axis.

@@

The presence of baryons produces rounder dark matter halos than found for the 
corresponding DMO counterpart. 
To better understand this relationship we examine the correlation
between halo shape and baryonic disc properties.  
Looking into the measurements already reported by \cite{auriga} and
\cite{Pakmor17}, we find three baryonic quantities that have the
strongest correlation with DM halo triaxiality: the central gas
density in a sphere of radius $1$ kpc, the disc to total mass ratio, and
the optical radius.

@@

In Figure~\ref{fig:twisting_correlations}, we show the correlation
between twisting and the same baryonic properties that showed a high
degree of correlation with the triaxiality.
In this case the disc to total ratio does not have a significant
degree of correlation. Instead, the disc optical radius and the gas density
at the core are the most correlated. 
The correlation is such that extended stellar discs with a very low
gas density at their cores show the most coherent halo shapes.

Recently, \citet{2017MNRAS.469..594B} used five zoom simulations 
to report that MW-like DM haloes can be roughly split into two
distinct families: stalled and accreting. 
Stalled haloes assembled early and have remained almost unchanged
in the last $7$ Gyr, while accreting halos have been growing
constantly. \citet{2017MNRAS.469..594B} also link the two populations to
differences in the dark matter filaments feeding the halos. 
Stalled halos at $z=0$ are embedded in a single large thick filament, while
accreting halos are pierced by many thinner filaments. 
To what extent can the twisting and high triaxiality present in the MHD
halos be linked to this dicotomy?

In Figure~\ref{fig:twisting_age} we show the correlation between the
twisting and the age of the bulge, the age of the stellar disc, and the
DM halo age.
We estimate the stellar ages  directly as the mean value in the
stellar age distribution for each component, while the
DM halo age corresponds to the lookback time at which the halo
consolidated half of its present mass.  
In the three cases we find a correlation between age and twisting
that goes in the direction claimed by \citet{2017MNRAS.469..594B}:
components with early assembly (i.e. stalled halos) have a smaller degree of
twisting. 
A detailed study to interpret this with respect to the cosmic 
web location of the halos is however beyond the scope of this paper.

@@

The second part of our findings correspond to the disc-halo alignment.
The dominant trend is that of a strong alignment between
the disc and the dark matter halo shape at all radii. 
In six halos of our sample we find a chaning alignment as a function
of radius.
This implies a twisting in the halo shape as a function of
radius.

Twisting has been also reported as an uncommon feature in some 
simulated systems found in the literature. 
\cite{JingSuto02} reported a shell alignment twisting in their dark
matter only simulations as measured in the direction changes of large
and medium shape axis. 
They only had three high resolution MW-like halos (with
$\sim 10^{6}$ inside the virial radius) and could not make a
statistical statement about the impact of this effect.
However, they see the twisting as an artifact resulting from high
values  of the $b/a$ ratios that make the determination of the medium
axis noisy, an effect that we have demonstrated can be discarded for
our simulations.
\cite{Bailin05} used seven hydrodynamic simulations to study the disc
halo alignment as a function of radius.
For radii smaller than $0.1R_{200}$ they find a strong global
alignment between the halo and the disc.
However, in some halos, at larger radii the angle between the
components twists resulting in a weaker correlation between the disk
and the halo shape. 
Considering that twisting is not present in our DMO simulations, it is
very likely that the detailed radial dependence of the alignments
between the disc and the halo depends on the hydrodynamics and
feedback implementations.

To finalize, we use the results reported by \cite{LM10} and \cite{Bovy16}
to place our results in an observational context.
\cite{LM10} used observations of the Sagittarius tidal stream to
constrain the shape of the gravitational potential.
They point out that previous studies that assumed an
axisymetric galactic potential were not able to fit all the available
dynamic constraints for the Sagittarius stream, motivating 
their approach of using a rigid triaxial potential with coaxial potential
ellipsoids for the dark matter component.  
Their results constrain the triaxiality of this potential
component.  They also translate their results into a triaxiality of the density
contours (which could be compared against our results)
 to be $c/a=0.44$ and $b/a=0.97$ at a radius of $\sim 40$~kpc. 
They do not report any uncertainties for these two values. 
Looking at their plots for the quality of fit criterion as a function
of dark halo axial scales (their Figure 5), we choose a conservative $10\%$
relative uncertainty. One surprising element in their results is that 
the major axis of the
halo shape is perpendicular to the stellar disc plane.

@@

\section*{Acknowledgements}
We acknowledge fruitful discussions with Patricia Tissera during the
preparation of the manuscript.
This project has received funding from the European Union's Horizon
2020 Research and Innovation Programme under the Marie
Sk\l{}odowska-Curie grant agreement No 734374.
